Speeding up eQTL scans in the BXD population using GPU

### 2019-07-25 paper.tex
BXD not BxD.
Added bibtex command to Makefile.
made small changes to manuscript

\title{Speeding up eQTL scans in the BxD population using GPU}

@@

The experiment setup is, to multiply two input matrices, A(mXn), and B(nXp), and produce an output matrix C(mXp)
The range of m, n and p is between 2\textsuperscript{4} and 2\textsuperscript{17} in log scale.
Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix total is less than 16 GB. 
The result is shown in Figure \ref{GPUCPUShape}. 
The X axis of Figure \ref{GPUCPUShape} is the dimensions of matrix whose I/O size is over 12 GB.
This is because using GPU involves a lot of overhead. 
Such overhead can only be justified by processing large data sets. 
The Y axis is the speedup of GPU comparing to CPU in log scale.  
From this figure, the matrices whose shape are closer to a square matrix, get better speedup from GPU.
Matrix multiplication is up to 5 times faster on GPU compared to 16 threaded CPU.

@@

\subsubsection{Single precision}
% Faster and memory efficient; may be less accurate, but acceptable for quick scan
Precision, in computer science, means the smallest difference between two representable numbers.
Floating point numbers, in scientific computing, are usually stored in double presicision. 
Double precision floating point numebrs takes up 8 bytes in memory, while single precision takes up 4 bytes. 
In addition to the difference in storing size, the speed for calculation using single and double precision also varies depending on hardware. 
For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS) for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80. 
Differences in optimization techniques, underlying architecture cause such performance gap. 
Considering the memory size, data transfer speed and throughput, single precision brings multiple benefits when precision is not the priority concern.

@@

To cope with this limitaion of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
Compared to CPU only, which took 0.55 seconds, this GPU modification provides 7 times speedup. 
For hippocampus datasets, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
As a backend service to a website, this wait time is much more preferable than the CPU timing for real time interactions.

>>>>

\title{Speeding up eQTL scans in the BXD population using GPU}

@@

The experiment setup is, to multiply two input matrices, A(m$\times$n), and B(n$\times$p), and produce an output matrix C(m$\times$p)
The range of m, n and p is between 2\textsuperscript{4} and 2\textsuperscript{17} in log scale.
Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix total is less than 16 GB. 
The result is shown in Figure \ref{GPUCPUShape}. 
The X axis of Figure \ref{GPUCPUShape} is the dimensions of matrix whose I/O size is over 12 GB.
This is because using GPU involves a lot of overhead. 
Such overhead can only be justified by processing large data sets. 
The Y axis is the speedup of GPU comparing to CPU in log scale.  
From this figure, the matrices whose shape are closer to a square matrix, get better speedup from GPU.
Matrix multiplication is up to 5 times faster on GPU compared to 16 threaded CPU.

@@

\subsubsection{Single precision}
% Faster and memory efficient; may be less accurate, but acceptable for quick scan
Precision means the smallest difference between two representable numbers.
Floating point numbers, in scientific computing, are usually stored in double presicision. 
Double precision floating point numebrs takes up 8 bytes in memory, while single precision takes up 4 bytes. 
In addition to the difference in storing size, the speed for calculation using single and double precision also varies depending on hardware. 
For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS) for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80. 
Differences in optimization techniques, underlying architecture cause such performance gap. 
Considering the storage size, data transfer speed and throughput, single precision brings multiple benefits when precision is not the priority concern.

@@

To cope with this limitaion of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
Compared to CPU only, which took 0.55 seconds, this GPU modification provides 7 times speedup. 
As we mentioned earlier, since we are porting the computation to GPU which took 90\% of total execution time, according to Amdahl's Law, the theoretical maximum speed up is 10 times. 
We got 7 times speed up for genome scan using GPU. 
For hippocampus datasets, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
As a backend service to a website, this wait time is much more preferable than the CPU timing for real time interactions.
